```{css, echo=FALSE}
.header-section-number::after {
  content: ".";
}
```

```{r}
# R Markdown template inside ~/.config/rstudio/templates
# set seed for reproducibility
set.seed(42)

# Load required library
packages <- c("tidyverse",
              "data.table",
              "janitor",
              "openxlsx",
              "RColorBrewer",
              "patchwork",
              "parallel",
              "sf",
              "tmaptools")

# load library quietly and stop if library can not be loaded
for (package in packages) {
  if (suppressPackageStartupMessages(require(package, character.only = TRUE))) {
  } else {
    stop("install required packages before running script")
  }
}

# source functions if defined in separate file
files = c("R/default.fun.R")
for (file in files) {
  if (file.exists(file)) {
    source(file)
  }
}
```



# World map with sf package

BACKGROUND: sf object is a special type of data frame with a geometry column that contains the spatial information of the object. The geometry column is a list of simple feature geometry objects. The sf object can be plotted using `geom_sf()` from ggplot2.

```{r map-world}
# function ne_countries() from rnaturalearth package
sf.world = rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")

as_tibble(sf.world)

# plot the world map with geom_sf
ggplot() +
  geom_sf(data = sf.world)
```

QUESTION: what is contain in 1 row of the `sf.world` object?

ANSWER: 1 row of the `sf.world` object contain the spatial information of a country. The geometry column contain the representation of the multi-polygon of the country (roughly speaking)

# Manipulate sf object

BACKGROUND: `sf` object can be manipulated using `dplyr` package. The `group_by()` function can be used to group the object by a column, and the `summarise()` function can be used to summarise the object. The `st_union()` function can be used to union the geometry of the object.

```{r}
# manipulate sf object ----
# group by continent
sf.continent = sf.world %>%
  st_make_valid() %>%
  group_by(continent) %>%
  summarise(geometry = st_union(geometry))

as_tibble(sf.continent)

# plot continent map with geom_sf
ggplot(sf.continent, aes(fill = continent)) +
  geom_sf() +
  theme(legend.position = "bottom")
# funny, all of russia is classified as europe # just live with it, that's the data we have
```

QUESTION: Any idea of why there are some distorted horizontal lines on the map?

ANSWER: The distorted horizontal lines are the results of a polygon crossing the date-line. So ggplot just draw a line across lon:-180:180

# Investigate the distorted polygon:

It cross the date-line at lon=180 or lon=-180

## Create sf object for a line with projection

NOTE: sf have various classes for spatial object, the most common are `sfc` and `sf`. `sfc` is a collection of simple feature geometry objects, while `sf` is a data frame with a `sfc` column and other columns that describe the attribute of the geometry.

THEY ARE NOT EQUIVALENT

QUESTION: what is the difference between: `sfg`, `sfc`, and `sf`?

ANSWER:

* `sfg` is a simple feature geometry object
* `sfc` is a list of simple feature geometry objects; with projection, bounding box INFO
* `sf` is a data frame with a `sfc` column and other columns that describe the attribute of the geometry. IMPORTANT: this is where `dplyr` function can be used to manipulate the object

```{r}
# create a sf line that have lon = 180 from matrix
sf.line = matrix(c(180, -90, 180, 90), ncol = 2, byrow = TRUE) %>%
  # convert into simple feature geometry # NOTE: `sfg`
  st_linestring() %>%
  # convert into simple feature geometry list column # NOTE: `sfc`
  st_sfc() %>%
  # convert into simple feature object # NOTE: `sf`
  st_sf(crs = 4326)
```

## Run sf operation to find which polygon cross the date line

There are a lot of sf operation that can be used to manipulate or check the spatial object.

READ MORE on all available geometry function in the cheatsheet of the `sf` package documentation: https://r-spatial.github.io/sf/

```{r}
# which polygon cross the date line
sf.cross = sf.world %>%
  # argument sparse = FALSE is required to make the function return a logical vector
  filter(st_crosses(geometry, sf.line,
                    sparse = FALSE)) %>%
  # select only field of interest # reminder that the column `geometry` will always be kept for object of class sf
  select(geounit, continent)

# print geounit and continent of the polygons that cross the date line
sf.cross %>% as_tibble()
```

## Plot the polygon before it is distorted

NOTE: this is the sf object, so we can use `geom_sf` to plot it.

You will see that we ignore aesthetic mapping in `aes()` entirely and just used the default options because it convenient

```{r}
ggplot(sf.cross) +
  geom_sf()
```

QUESTION


# How do we fix the distorted polygon

Break up the polygon that cross the date line with `st_break_antimeridian` function

```{r map-fix, message=FALSE, warning=FALSE}
# manipulate sf object ----
# THIS IS THE FIX FOR ST_BREAK_ANTIMERIDIAN
# turn off the google s2geometry.io library # and use the GEOS library
sf_use_s2(FALSE)
# group by continent
sf.continent = sf.world %>%
  # this is the fix for st_union ----
  st_make_valid() %>%
  group_by(continent) %>%
  summarise(geometry = st_union(geometry)) %>%
  # this is the important fix for distorted polygon that cross the date line ----
  st_break_antimeridian(lon_0 = 0)

# plot continent map with geom_sf
fig = ggplot(sf.continent, aes(fill = continent)) +
  geom_sf() +
  # don't forget that this is still a ggplot object than can be modified with theme
  theme(legend.position = "bottom",
        panel.background = element_rect(fill = "lightskyblue"))
fig
```

# Add geom_point layer to the map (ggplot)

Creating sf object for the data is too much WORK. We can just use a data frame with x and y coordinate and use `geom_point` to plot it.

```{r}
# generate random points with x and y coordinate
# it is saved inside a data frame
df.coord = tibble(x = runif(10, -180, 180),
                   y = runif(10, -90, 90))
df.coord
# add geom_point layer to the map
fig2 = fig +
  geom_point(data = df.coord,
             aes(x = x, y = y),
             # this is important, make sure this layer don't inherit the aes from the previous layer
             inherit.aes = FALSE,
             color = "red", size = 3) +
  # do remember that `df.coord` have no projection, so we need to specify it
  # easiest way to do is to use coord_sf(default_crs = 4326) # just use the default 4326 for any coordinate without projection crs
  coord_sf(default_crs = 4326)
fig2
```

# Change the projection of the map

I like the Robinson projection, so let's do this with `coord_sf`

This can be a little complicated with the way map string work.

Take a little bit of time and google to get the correct string for the projection you want.

## Robinson projection

```{r}
# declare the string for the robinson projection
robinson_str = "+proj=robin +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs"

fig2 +
  coord_sf(crs = robinson_str,
           default_crs = 4326)
```

## Fixing the background

LOOK: the map work, but the background do not fit the area of water anymore, because we use `theme(panel.background)`, which always change the color of the rectangle area.

BUT: we want to color the oval area of the map.

QUESTION: how do we fix this?

ANSWER: The idea is add ANOTHER layer to represent the projected map: make an oval shape with `geom_sf` and color it.

```{r}
# there is a function to create a oval shape in the `tmaptools` package
sf.ocean = tmaptools::bb_earth(projection = "WGS84",
  stepsize = 1,
  earth.datum = 4326) %>%
  st_as_sf()

# plot the oval shape as the 1st layer
fig.rob = ggplot() +
  # we need alpha = 0.5 because otherwise we cannot see the  lon-lat grid on the panel layer
  geom_sf(data = sf.ocean, fill = "lightskyblue", alpha = 0.5) +
  # add the sf.continent layer
  geom_sf(data = sf.continent, aes(fill = continent)) +
  # add the geom_point layer
  geom_point(data = df.coord,
             aes(x = x, y = y),
             inherit.aes = FALSE,
             color = "red", size = 3) +
  # change the projection
  coord_sf(crs = robinson_str,
           default_crs = 4326) +
  # theme for ggplot object
  theme(legend.position = "bottom",
        panel.background = element_blank(),
        panel.grid = element_line(color = "grey"),
        axis.title = element_blank())
fig.rob
```

## zoom in to Australia and change to conic projection

```{r}
# proj4 string for the conic projection above Australia
conic_str = '+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-25 +lon_0=135 +ellps=aust_SA  +units=m +no_defs'

# bounding box of australia extracted from sf object of countries
bbox = sf.world %>%
  filter(geounit == "Australia") %>%
  st_bbox()
bbox
# plot the map with the new projection
fig.rob +
  # change the projection
  coord_sf(crs = conic_str,
           default_crs = 4326,
           xlim = c(bbox$xmin, bbox$xmax), ylim = c(bbox$ymin, bbox$ymax))
```

LOOK at the shape of the longitude and latitude grid, it is CONIC projection. YAY!

# Idea for tutorial

Extract Singapore from sf.world and plot it on the map with `mercator` projection

```{r}

```


